---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# statnnet

`statnnet` provides regression-style statistical summaries for small,
single-hidden-layer neural networks fitted with `nnet`. It retains the
original fit: it is not a neural-network optimiser or model-selection package.

## Installation

``` r
# install.packages("pak")
pak::pak("andrew-mcinerney/statnnet")
```

## Basic workflow

```{r example, message=FALSE, warning=FALSE}
library(nnet)
library(statnnet)

set.seed(2026)
fit <- nnet(
  medv ~ rm + lstat + chas,
  data = Boston,
  size = 1,
  linout = TRUE,
  decay = 0.1,
  Hess = TRUE,
  maxit = 1000,
  trace = FALSE
)

model <- statnnet(
  fit,
  formula = medv ~ rm + lstat + chas,
  data = Boston,
  response = "continuous"
)

model
anova(model)
```

`summary(model)` returns structured average partial covariate effects and
grouped Wald-type summaries. `pce()` and `plot()` provide finite-difference
partial covariate effects with delta-method or simulation uncertainty.

```{r pce, eval=FALSE}
summary(model)
pce(model, "rm")
plot(model, "rm")
plotnn(model, annotate = "p_value")
```

These summaries describe the fitted response function. They are conditional
on the chosen network and are not post-selection or causal inference.
