---
title: "A statnnet workflow with nnet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A statnnet workflow with nnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`statnnet` starts from a fitted `nnet` model and does not refit it. This example
uses a small linear-output network so that the covariance diagnostics and
regression-style summaries remain interpretable.

```{r setup, message=FALSE}
library(nnet)
library(statnnet)

data(Boston)
set.seed(2026)
fit <- nnet(
  medv ~ rm + lstat + chas,
  data = Boston,
  size = 1,
  linout = TRUE,
  decay = 0.1,
  Hess = TRUE,
  maxit = 1000,
  trace = FALSE
)
```

The ordinary `nnet` summary remains available on the original fit. The
`statnnet` object adds validation, covariance diagnostics, grouped tests, and
two explicitly distinguished finite-difference effect summaries.

```{r augment}
summary(fit)

model <- statnnet(
  fit,
  formula = medv ~ rm + lstat + chas,
  data = Boston,
  response = "continuous"
)

model
summary(model, effects = "rowwise")
```

The default `"rowwise"` summary changes the focal predictor within each
observed covariate row before averaging the prediction differences. It is not
an average partial-dependence effect. The more computationally intensive
average partial-dependence contrast is requested explicitly:

```{r partial-summary}
summary(model, effects = "partial")
```

`anova()` groups all model-matrix columns belonging to the same formula term.
This is particularly important when a factor expands to multiple dummy
variables.

```{r grouped}
anova(model)
```

`pce()` returns finite differences of the partial-dependence function rather
than the default observed-row summary. These effects describe the fitted
response function and should not be interpreted causally. Correlated
predictors may lead to poorly supported predictor combinations.

```{r pce}
pce(model, "rm", length_out = 5)
```
